---
title: "ASM Correlations"
format: html
editor: visual
---

## Correlations for ASM

Mackenzi had asked me to look into associations between the top 10 anti-epileptic meds and individuals who have only a functional seizures, and those who have a dual diagnosis of functional seizures AND epilepsy.

Furthermore, we want to look into specific correlations for individuals who have comorbid: mood disorders, Migraines, or Chronic pain.

## Packages:

```{r}
library(tidyverse)
library(readxl)
library(corrplot)
library(epitools)
```

## Data Import

```{r}

# Define the file path
file_path <- "C:\\Users\\domin\\Downloads\\ASM.xlsx"

# Read the first sheet
asm <- read_excel(file_path, sheet = 1) 
```

## Top 10 Meds:

Our list of the most common meds are:

-   Levetiracetam

-   Topiramate

-   Sodium Valproate

-   Lamotrigine

-   Gabapentin

-   Oxcarbazepine

-   Zonisamide

-   Phenytoin

-   Lacosamide

-   Carbamazepine

# Correlations

I want to start with broad strokes of correlations and then increasing become more nuanced. Let's start with just the medications and dual diagnosis:

## Meds + Dual Dx

Let's compare those 10 meds with our binary outcome variable Dual_Dx

```{r}

# First, let's change Dual_Dx to a number:
asm$Dual_Dx <- ifelse(asm$Dual_Dx == "Yes", 1, 0)


# Loop through all medication variables and run Chi-Square tests
for (med in c("Med_Levetiracetam", "Med_Topiramate", "Med_Sodium Valproate", "Med_Lamotrigine", "Med_Gabapentin", "Med_Oxcarbazepine", "Med_Zonisamide", "Med_Phenytoin", "Med_Lacosamide", "Med_Carbamazepine")) {
  tab <- table(asm$Dual_Dx, asm[[med]])  # Create a contingency table
  chi_test <- chisq.test(tab)  # Perform Chi-Square test
  cat("\nChi-Square Test for", med, "\n")
  print(chi_test)
}

```

Great, we have significant results for:

-   Sodium Valproate (p = 1.156e-05)

-   Gabapentin (p = .02302)

-   Zonisamind (p = .001874)

-   Phenytouin ( p = 5.205e-09)

-   Lacosamide (p = .01844)

-   Carbamazepine (p = 4.921e-07)

Now we need to examine direction to make causal statements:

### Assessing Directionality

```{r}

# Let's look at the Sodium Valproate first:
table(asm$Dual_Dx, asm$`Med_Sodium Valproate`)

# Odds ratio for strength of association:
or_sodiumVal <- oddsratio(table(asm$Dual_Dx, asm$`Med_Sodium Valproate`))
print(or_sodiumVal$measure)

# On average, people with a dual dx are 2.53 times more likely to use Sodium Valproate than those without a dual diagnosis.

```

Great, this tracks. Individuals who have a separate diagnosis of epilepsy tend to have a 2.5 times higher rate of use of sodium valproate.

However, we do have 4 medications that are not statistically significant in the chi-square test. This strikes me as strange because why would there NOT be a difference in these groups. Especially if one of the groups does not have an epilepsy diagnosis.

Let's look at topiramate for example:

```{r}
table(asm$Dual_Dx, asm$Med_Topiramate)

# Odds ratio for strength of association:
or_topiramte <- oddsratio(table(asm$Dual_Dx, asm$Med_Topiramate))
print(or_topiramte$measure)

# On average, those with a dual dx are .09 times more likely to use topiramate. not clinically significant. 
```

Makes sense, so these groups are receiving a epilepsy drug at roughly the same rate as a those with an epilepsy dx. That's interesting.

## Adding in MedHx:

Ok, now that we've built the framework, let's move on to further splitting these groups into their respective categories based on comorbidities.

We'll make 3 separate databases, one for those with mood disorders, one for migraine, and one for chronic pain.

### Mood disorder:

Mood disorders are classified as any of the following:

-   GAD, Anxiety, depression, major depressive disorder, bipolar disorder, and mood disorder.

Let's create this variable and create a new dataframe:

```{r}

# Let's start by making our new binary variable for mood disorders:

asm <- asm %>%
  mutate(Mood_Disorder = ifelse(Hist_Depression == 1 | `Hist_Generalized Anxiety` == 1 | Hist_Anxiety == 1 | `Hist_Major depressive disorder` == 1 | Hist_Bipolar == 1 | `Hist_Mood disorder` == 1, 1, 0))

# Now let's filter the database to only have patients with a mood disorder

asm_mood <- asm %>% filter(Mood_Disorder == 1)


```

Great, we have a working database, now let's see run our previous code on this one and see our results:

```{r}
# Loop through all medication variables and run Chi-Square tests
for (med in c("Med_Levetiracetam", "Med_Topiramate", "Med_Sodium Valproate", "Med_Lamotrigine", "Med_Gabapentin", "Med_Oxcarbazepine", "Med_Zonisamide", "Med_Phenytoin", "Med_Lacosamide", "Med_Carbamazepine")) {
  tab <- table(asm_mood$Dual_Dx, asm_mood[[med]])  # Create a contingency table
  chi_test <- chisq.test(tab)  # Perform Chi-Square test
  cat("\nChi-Square Test for", med, "\n")
  print(chi_test)
}
```

Significant results:

Sodium Valproate, Zonisamide, Phenytoin, Lacosamide, Cabamazepine:

Lamotragine and Oxcarbazepine are approaching significance. Let's run our two tests and see how this fits:

```{r}

# Let's run this with Sodium valproate as an example

table(asm_mood$Dual_Dx, asm_mood$`Med_Sodium Valproate`)

# Odds ratio for strength of association:
or_sodium_mood <- oddsratio(table(asm_mood$Dual_Dx, asm_mood$`Med_Sodium Valproate`))
print(or_sodium_mood$measure)

# Individuals with a dual diagnosis are 2.5 times more likely to take sodium valproate.
```

Let's do the same but for a non-significant value: Lamotragine:

```{r}
table(asm_mood$Dual_Dx, asm_mood$Med_Lamotrigine)

# Odds ratio for strength of association:
or_lamotragine_mood <- oddsratio(table(asm_mood$Dual_Dx, asm_mood$Med_Lamotrigine))
print(or_lamotragine_mood$measure)

# Individuals who have a dual dx are .61 times more likely to take lamotrigine than those who do not have a dual dx. Not clinically significant.
```

Similar story with mood disorders. Let's break this down to the other two groups and see if the story is the same.

### Migraine:

```{r}
# Let's start by making our new binary variable for migraine:

asm <- asm %>%
  mutate(migraine = ifelse(Hist_Migraine == 1 | Hist_Headache == 1, 1, 0))

# Now let's filter the database to only have patients with a mood disorder

asm_migraine <- asm %>% filter(migraine == 1)
```

Same as before, let's run our function to assess chi-square values:

```{r}
# Loop through all medication variables and run Chi-Square tests
for (med in c("Med_Levetiracetam", "Med_Topiramate", "Med_Sodium Valproate", "Med_Lamotrigine", "Med_Gabapentin", "Med_Oxcarbazepine", "Med_Zonisamide", "Med_Phenytoin", "Med_Lacosamide", "Med_Carbamazepine")) {
  tab <- table(asm_migraine$Dual_Dx, asm_migraine[[med]])  # Create a contingency table
  chi_test <- chisq.test(tab)  # Perform Chi-Square test
  cat("\nChi-Square Test for", med, "\n")
  print(chi_test)
}
```

Significant results:

-   Sodium Valproate, Zonisamide, Phenytoin, Lacosamide, Carbamazepine

Same story: let's look at odds ratios to determine direction:

```{r}

table(asm_migraine$Dual_Dx, asm_migraine$`Med_Sodium Valproate`)

# Odds ratio for strength of association:
or_sodium_migraine <- oddsratio(table(asm_migraine$Dual_Dx, asm_migraine$`Med_Sodium Valproate`))
print(or_sodium_migraine$measure)

#Individuals with a dual dx are 3.2 times more likey to use sodium valproate on average. Statistically significant.
```

Now with a non-significant:

```{r}

table(asm_migraine$Dual_Dx, asm_migraine$Med_Lamotrigine)

# Odds ratio for strength of association:
or_lamotrigine_migraine <- oddsratio(table(asm_migraine$Dual_Dx, asm_migraine$Med_Lamotrigine))
print(or_lamotrigine_migraine$measure)

# Individuals with a dual dx are 1.69 times more likely to take lamotrigine when compared to those without a dual dx. not clinically significant.
```

Same story. Let's check chronic pain just in case.

### Chronic Pain:

```{r}
# Let's start by making our new binary variable for migraine:

asm <- asm %>%
  mutate(chronic = ifelse(`Hist_Chronic Pain` == 1, 1, 0))

# Now let's filter the database to only have patients with a mood disorder

asm_chronic <- asm %>% filter(chronic == 1)
```

Same as before, let's run our function for chronic pain:

```{r}
# Loop through all medication variables and run Chi-Square tests
for (med in c("Med_Levetiracetam", "Med_Topiramate", "Med_Sodium Valproate", "Med_Lamotrigine", "Med_Gabapentin", "Med_Oxcarbazepine", "Med_Zonisamide", "Med_Phenytoin", "Med_Lacosamide", "Med_Carbamazepine")) {
  tab <- table(asm_chronic$Dual_Dx, asm_chronic[[med]])  # Create a contingency table
  chi_test <- chisq.test(tab)  # Perform Chi-Square test
  cat("\nChi-Square Test for", med, "\n")
  print(chi_test)
}
```

Statistically significant:

-   Sodium Valproate, Carbamazepine (low N),

Same as before, let's check directionality:

```{r}
table(asm_chronic$Dual_Dx, asm_chronic$`Med_Sodium Valproate`)

# Odds ratio for strength of association:
or_sodium_chronic <- oddsratio(table(asm_chronic$Dual_Dx, asm_chronic$`Med_Sodium Valproate`))
print(or_sodium_chronic$measure)

# Individuals with a dual diagnosis are 2.83 times more likely to be prescribed sodium valproate when compared to individuals without a dual dx. Stat sig.
```

Now for non-significant

```{r}
table(asm_chronic$Dual_Dx, asm_chronic$Med_Lamotrigine)

# Odds ratio for strength of association:
or_lamotrigine_chronic <- oddsratio(table(asm_chronic$Dual_Dx, asm_chronic$Med_Lamotrigine))
print(or_lamotrigine_chronic$measure)

# Individuals with a dual diagnosis are 2.68 times more likely to use lamotrigine when compared to those without a dual dx. Not clinicall significant.
```
